---
title: "Portfolio Template (Note - not all of these plots are A-worthy plots!)"
output: pdf_document
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, results = 'hide', warning = FALSE, message=FALSE)
```
# Reflection
*The following is an example reflection page. You don't need to format yours exactly like this, but you can if you like. Please do NOT use the embedded R datasets, I've done so only for this example but yours should use real, external data either from the course or that you found on your own.*
In this portfolio I have created 4 plots and 1 table to show how I have mastered the criteria listed in the syllabus. Throughout the semester your feedback has indicated I needed to work particularly hard on a few criteria that I was initially weak on. I struggled at first to master #4, but you'll see that in plot 4 in this portfolio I noticed, addressed, and described a serious data quality issue that would have otherwise compromised the plot. I was also struggling at first with number 6 and 7, receiving the feedback that I was over-complicating plots in a way that made them difficult to understand the answer to the question. I've worked hard at addressing these points in plot 2 and plot 3.

In most assignments I made one or two minor errors or had a chart illusion I didn't catch. With practice, I have now successfully eliminated these problems (criteria 12 & 13) in all of the plots that follow (I hope, anyhow). As a result, I think I deserve an **A** in the course, as my portfolio shows I have mastered almost all the criteria and have improved dramatically on those that I was struggling with before. 

My portfolio plots are designed to specifically show mastery in the following criteria:

*  Whole Portfolio - 2,3,10,11,12,13,14
*  1, a plot - 1,5,7
*  2, a table - 8, 9, 4 
*  3, a faceted plot - 6, 7
*  4, a basic bar plot, 7
*  5, a multi-paneled plot - 5, 6

\newpage
```{r plot1}
library(dplyr)
library(reshape2)
library(ggplot2)
library(palmerpenguins) #so I can use this dataset. 
#You shouldn't though! Use a bigger one, not the in-package ones.

#If you are using datasets from outside the class, you could provide links 
#to them here, or in the discord, or in the text if you can 
#find a good place to fit it. 

#Code to make your first plot here, 
#which should be more exciting than the one I'm demoing.

ggplot(iris, aes(x = Sepal.Length, fill = Species)) + 
  geom_density(alpha = .5) + 
  labs(x = "Sepal Length (mm)", y = "Density")+
  scale_fill_manual(values = c("firebrick", "salmon", "pink"),
                    name = "Species",
                    breaks = c("setosa", "versicolor", "virginica"),
                    labels = c("I. setosa", "I. versicolor", "I. virginica"))+
  theme_classic()
```
**Plot 1** - this is where your caption goes. Make sure to describe the figure, give important background on the data, and give a little interpretation of the trend you see. You can also say what question you were trying to answer, since you may have made one up on your own. If you really want to drill home mastery of the syllabus criteria, you can briefly mention which criteria this plot addresses.So for example: *This density estimate plot shows the differences in the length of the sepal of three species of iris. I. setosa is the smallest, while I. virginica is the largest. There are many other species of iris; these are the only three included in the embedded iris dataset.*

\newpage
```{r plot2, results = "show"}
olympic <- read.csv("Datasets/olympic athlete_events.csv")
o2 <- olympic %>%
  filter(Season == "Summer") %>%
  filter (Year >=1950) %>%
  filter(Medal == "Gold") %>%
  group_by(Name, Event, Team, Sex, Year) %>%
  tally() %>%
  group_by(Year, Team) %>%
  tally() %>%
  filter(Team %in% c("China", "United States", "Russia")) %>%
  dcast(Year ~ Team, value.var = "n", fill = 0)

knitr::kable(o2,
             booktabs = T,
             align = "crrr")
```
**Table 1** - The number of gold medals brought home by three countries for each summer Olympics since the games resumed post World War II. The United States has continued to bring home quite a few. 


*Table Notes:*

*  Tables do sometimes have notes. This one doesn't really, so hence why these are... the way they are. 
*  These should be bullet points, and separated from your figure
*  Make them very brief!

\newpage
```{r plot3}
pengmelt <- penguins %>%
  melt(id.var = c("species", "sex", "island", "year", "body_mass_g"))
ggplot(pengmelt, aes(x = body_mass_g, col = species, y = value)) + 
  geom_point() + 
  facet_wrap(~variable, scales = "free")+
  labs(x = "Body Mass (g)", y = "Measurement")+
  scale_color_manual(values = c("red", "blue", "darkgreen"))+
  theme_bw()
```
**Plot 3** - A facetted scatterplot that shows how different characteristics split out the three different species of penguin in the Palmer Penguins dataset. Bill Depth by body mass shows a clear separation for one species, but Bill Length by body mass shows separation of all three.

\newpage
```{r plot4}
ggplot(penguins, aes(x = species))+ geom_bar()+
  labs(x = "Species", y = "Count")
```
**Plot 4** - Not all plots need to be complicated. While this is not as colorful as the others, this barplot shows the total number of observations in the embedded penguin dataset by species in a way that is clear and minimizes cognitive load (#7). 

\newpage
```{r plot5}
library(gridExtra)
g1 <- ggplot(penguins %>% filter(!is.na(sex)), aes(x = flipper_length_mm, fill = sex))+
  geom_histogram(position = "dodge")+
  labs(x = "Flipper Length (mm) by Sex", y = "Count")+
  theme(legend.position = "top")+
        annotate(geom = 'text', label = 'A ', x = Inf, y = -Inf, 
                 hjust = 1, vjust = -1, size =14)

g2 <- ggplot(penguins, aes(x = flipper_length_mm, fill = species))+
  geom_histogram(position = "dodge")+
  scale_fill_manual(values = c("dodgerblue", "cornflowerblue", "blue"))+
  labs(x = "Flipper Length (mm) by Species", y = "Count")+
  theme(legend.position = "bottom")+
        annotate(geom = 'text', label = 'B ', x = Inf, y = -Inf, 
                 hjust = 1, vjust = -1, size =14)

grid.arrange(g1,g2)
```
**Plot 5** - This mult-paneled dodged histogram shows how sex (Panel A) and species (Panel B) can influence flipper length in penguins. Females typically have smaller flippers, but so do Adelie Penguins as a whole. This data is from the Palmer Penguin dataset. 

\newpage
# Code Output

```{r plot1, eval = FALSE, echo = TRUE, fig.show = "hide"}
```

```{r plot2, eval = FALSE, echo = TRUE, fig.show = "hide"}
```

```{r plot3, eval = FALSE, echo = TRUE, fig.show = "hide"}
```

```{r plot4, eval = FALSE, echo = TRUE, fig.show = "hide"}
```

```{r plot5, eval = FALSE, echo = TRUE, fig.show = "hide"}
```
